### 2015-01-13 doc/thesis/system.tex
Starting the presentation of typing rules with assignments

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $c$ are literal constants, $p$ are expression pairs,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $n$ are variable names,
$t$ are first-level types, and $r$ are second-level types.
The notation $\vec{n{:}t}$ denotes the non-empty list
$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.

@@

We will present the typing rules as a deduction system for the typing
relations $\env_{1} \vdash s : \env_{2}$ and $\env_{1} \vdash e : t, \env_{2}$. 
We will use the first one for typing statements, and the second one
for typing expressions.
The first relation means that given the type environment $\env_{1}$,
typing the statement $s$ produces the new type environment $\env_{2}$. 
The second relation means that given the type environment $\env_{1}$,
the expression $e$ has type $t$, and produces the new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though both expressions and statements can modify the environment.
The environment maps variables to types, and we will use $\env_{1}[n \mapsto t]$
for meaning that we extended the environment $\env_{1}$ with the
variable $n$ that maps to type $t$, and this mapping does not affect
the other mappings that existed before.

@@

Lua has multiple assignments, so Typed Lua also has to take this into
consideration while refining table types.
The rule \textsc{T-ASSINGMENT} shows that tuple types make it easy
to use the consistent-subtyping relation to type check multiple assignments:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \vec{l} = el:\env_{3}}
\end{array}
\end{array}
\]

This rule is so simple because we left the difficult part to the
type checking of expression lists.
The list of left-hand values is also a kind of expression list.
The difficult part is to check whether there is an expression in
the expression list that messes the environment, and the rule
\textsc{T-LHSLIST} captures this intuition:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env \vdash l_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
to the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.

>>>>

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $c$ are literal constants, $p$ are table fields,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $n$ are variable names,
$t$ are first-level types, and $r$ are second-level types.
The notation $\vec{n{:}t}$ denotes the non-empty list
$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.

@@

We present the typing rules as a deduction system for the typing
relations $\env_{1}, \penv \vdash s, \env_{2}$ and
$\env_{1}, \penv \vdash e : t, \env_{2}$.
We use the first one for typing statements, and the second one
for typing expressions.
The first relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
typing the statement $s$ produces the new type environment $\env_{2}$.
The second relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify the environment.
We need two environments because $\env$ maps variables to types
while $\penv$ maps projection variables to unions of tuple types.
We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
$\env_{1}$ with the variable $n$ that maps to type $t$.
We use $\penv[\pi \mapsto r]$ for meaning that we extendend the environment
$\penv$ with the projection $\pi$ that maps to the union of tuple types $r$.

Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consitent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \vec{l} = el:\env_{3}}
\end{array}
\end{array}
\]

As an example,
\[
x, y = 1, ``foo"
\]
type checks because
$1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}$,
assuming that $x$ and $y$ are in the environment with
types $\Integer$ and $\String$, respectively.
Our type system uses $\Value{*}$ to drop extra values
and $\Nil{*}$ to replace missing values.
In this example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
are the rules that handle these features.

We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env \vdash l_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
to the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.
We will show in this section that the refinment of table types can
change the environment, and that is the reason why we need to check
whether the environment is consistent.

The rule \textsc{T-EXPLIST2} is close to the rule \textsc{T-LHSLIST}.
The difference is that it uses $\Nil{*}$ to replace missing values.
We define this rule as follows:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

The typing rules that type check expression lists include $\Nil{*}$
in the end of the tuple type only when it does not end in a variadic type.
For instance, the rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4}
handle the case that an expression list is just an expression that
return multiple values, but only \textsc{T-EXPLIST3} needs to add
$\Nil{*}$:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{1}}
      {\env \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{1}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
      {\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
\end{array}
\end{array}
\]

As an example, the rule \textsc{T-EXPLIST3} allows our type system
to type check the following example:
\[
x, y, z = f()_{m}
\]
assuming that $x$, $y$, $z$, and $f$ are in the environment with
types $\Integer$, $\String$, $\String \cup \Nil$, and
$\Void \rightarrow \Integer \times \String \times \Void$, respectively.
It type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

Conversely,
\[
x = f()_{m}
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
\]

Multiple assignments also appear in function and method applications,
but our type system needs different typing rules for
expression lists that appear in function and method applications,
as we also want to catch arity mismatch.
For instance, assuming that $x$ and $f$ are in the environment with
types $\Integer \cup \Nil$ and
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
respectively.
We want that
\[
x = f(``foo")_{m}
\]
type checks because
\[
``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but we also want that
\[
x = f(``foo",1,2,3)_{m}
\]
does not type check because
\[
``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

To do that, the rule \textsc{T-APPLY1} uses rules that type check
expression lists, but that do not include $\Nil{*}$ in the end of
the tuple type.
Instead, it uses the predicate $adjust$ to replace any missing
value with $\Nil$:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
       r_{4} = adjust(r_{3}, r_{1}) \;\;\;
       r_{4} \lesssim r_{1}}
      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

We can use pattern matching to the define the recursive predicate
\emph{adjust} as follows:
\begin{align*}
adjust(\Void, t{*}) & = \Nil \times \Void\\
adjust(\Void, t \times s) & = \Nil \times adjust(\Void,s)\\
adjust(t_{1} \times s_{1}, t_{2} \times s_{2}) & = t_{1} \times adjust(s1,s2)\\
adjust(s_{1}, s_{2}) & = s_{1}
\end{align*}

@@
